EDA of Red Wine Quality Dataset by Iwan Thomas
========================================================

```{r global_options, include=FALSE}
knitr::opts_chunk$set(echo=FALSE, warning=FALSE, message=FALSE)
```

```{r echo=FALSE, message=FALSE, warning=FALSE, results='hide', packages}
# Load all of the packages that you end up using
# in your analysis in this code chunk.

# Notice that the parameter "echo" was set to FALSE for this code chunk.
# This prevents the code from displaying in the knitted HTML output.
# You should set echo=FALSE for all code chunks in your file.

install.packages("gridExtra", repos = "https://cran.ma.imperial.ac.uk/")
install.packages("ggplot2", repos = "https://cran.ma.imperial.ac.uk/")
install.packages("GGally", repos = "https://cran.ma.imperial.ac.uk/")

library(gridExtra)
library(ggplot2)
library(GGally)
library(RColorBrewer)
```

# Introduction
In this analysis, I will explore a dataset of 1599 individual red wines with the aim of determining what chemical properties influence their quality. I will perform this analysis in R. Each individual wine is described by:

* a unique identifier

* 11 chemical properties  

* an expert quality score, where 0 is very bad and 10 is very excellent.

The data used was from the following source:

* P. Cortez, A. Cerdeira, F. Almeida, T. Matos and J. Reis.

* Modeling wine preferences by data mining from physicochemical properties.

* In Decision Support Systems, Elsevier, 47(4):547-553. ISSN: 0167-9236.

The original dataset can be found [here](https://www.google.com/url?q=https://s3.amazonaws.com/udacity-hosted-downloads/ud651/wineQualityReds.csv&sa=D&ust=1472652941026000&usg=AFQjCNH7NX-Zos0LE3pSojCeMI66wf7j3Q) and a description of the variables and how the data was collected can be found [here](https://s3.amazonaws.com/udacity-hosted-downloads/ud651/wineQualityInfo.txt). 

I will first load and inspect the dataset. 

```{r echo=FALSE, Load_the_Data}
# Load the Data

setwd("C:/Users/IThomas/Documents/Udacity DAND/P4/Final Project")
wine <- read.csv("wineQualityReds.csv")
str(wine)
summary(wine)

```

The data is tidy and clean. The only surprise I noted was 132 wines with citric acid values of zero. 

Let's first look at the distributions of all of our properties. Some of the plots are right-skewed and a logarithmic transformation of the x-axis was applied to further explore the variable's distribution.

* Quality, alcohol, volatile acidity, citric acid (when the anomalous zero readings are discounted), density, pH and fixed acidity are approximately normally distributed. 
 
* The x-axes of the variables chlorides, residual sugar, total sulfur dioxide and free sulfur dioxide were log transformed. 


# Univariate Plots Section
```{r echo=FALSE, Univariate_Plots_1}
univ_plot <- function(variable, bins) {
  ggplot(aes_string(x = variable), data = wine) + 
    geom_histogram(binwidth = bins)
}

p0 <- univ_plot("quality", 1)
p1 <- univ_plot("alcohol", 0.2)
p2 <- univ_plot("volatile.acidity", 0.05)
p3 <- univ_plot("sulphates", 0.04)
p4 <- univ_plot("citric.acid", 0.05)
p5 <- univ_plot("total.sulfur.dioxide", 5)
p6 <- univ_plot("density", 0.0005)
p7 <- univ_plot("chlorides", 0.005)
p8 <- univ_plot("pH", 0.05)
p9 <- univ_plot("free.sulfur.dioxide", 2)
p10 <- univ_plot("residual.sugar", 0.1)
p11 <- univ_plot("fixed.acidity", 0.5)

grid.arrange(p0, p1, p2, p3, p4, p5, p6, p7, p8, p9, p10, p11, ncol = 3)
```

Let's replot this figure with a logarithmic transformation of the x-axis of some variables, namely chlorides, residual sugar, total sulfur dioxide and free sulfur dioxide.

```{r echo=FALSE, Univariate_Plots_2}

p5_log <- univ_plot("total.sulfur.dioxide", .05) + scale_x_log10()
p7_log <- univ_plot("chlorides", 0.02) + scale_x_log10()
p9_log <- univ_plot("free.sulfur.dioxide", 0.1) + scale_x_log10()
p10_log <- univ_plot("residual.sugar", 0.05) + scale_x_log10()

grid.arrange(p0, p1, p2, p3, p4, p5_log, p6, p7_log, p8, p9_log, p10_log, p11, ncol = 3)

summary(wine)

# Some of these plots are right-skewed and warrant further exploration with a logarithmic transformation of the x-axis. 

# Quality, alcohol, volatile acidity, citric acid (when the anomalous zero readings are discounted), density, pH and fixed acidity are approximately normally distributed. 

# Let's investigate the other properties when the x-axis is transformed. 

```

Let's look at some of these distributions in greater detail. Relevant statistics are included at the end of each plot. 

```{r echo=FALSE, Univariate_Plots_3}
qplot(x = quality, 
      data = wine, 
      binwidth = 1,
      color = I('black'), 
      fill = I('#099DD9')) + 
  scale_x_continuous(breaks = seq(2,8,1))

summary(wine$quality)

grid.arrange(ggplot(wine, aes( x = 1, y = alcohol) ) + 
               geom_boxplot( ) +
               geom_point(alpha = 0.1, position = "jitter") 
             ,
             ggplot(wine, aes( x   = alcohol  ) ) + 
                   geom_histogram(bins=30 ),ncol=2)

summary(wine$alcohol)

grid.arrange(ggplot(wine, aes( x = 1, y = volatile.acidity ) ) + 
               geom_boxplot( ) +
               geom_jitter(alpha = 0.1 ) 
             ,
             ggplot(wine, aes( x   = volatile.acidity  ) ) + 
                   geom_histogram(bins=30 ),ncol=2)

summary(wine$volatile.acidity)

grid.arrange(ggplot(wine, aes( x = 1, y = citric.acid ) ) + 
               geom_boxplot( ) +
               geom_jitter(alpha = 0.1 ) 
             ,
             ggplot(wine, aes( x   = citric.acid  ) ) + 
                   geom_histogram(bins=30 ),ncol=2)

summary(wine$citric.acid)

grid.arrange(ggplot(wine, aes( x = 1, y = sulphates ) ) + 
               geom_boxplot( ) + 
               geom_jitter(alpha = 0.1 )
             ,
             ggplot(wine, aes( x   = sulphates  ) ) + 
                   geom_histogram(bins=30 ),ncol=2)

summary(wine$sulphates)
```

* We see that wine quality is approximately normally distributed with a mean value of 5.6.

* The alcohol distribution is right-skewed. It will be interesting to see whether different quality wines have distinct alcohol content. 

* Volatile acidity has quite a few outliers with high values. Which wines do these values belong to?

* I have doubts about the validity of the citric acid data, because of the numbers of wines with citric acid = 0. 

* The sulphates distribution is quite tight with a mean of 0.66. I'm interested in understanding whether the outlying data point belong to good, average or bad wines. 


# Univariate Analysis

### What is the structure of your dataset?
There are 1599 observations of 13 variables. Each wine is assigned an integer id value, 11  numerical chemical properties of wine are recorded, and the wine is awarded a quality rating.

### What is/are the main feature(s) of interest in your dataset?
The main feature of interest is the quality rating. I will investigate the relationship between various chemical properties and the quality of the wine. 

### What other features in the dataset do you think will help support your investigation into your feature(s) of interest?
Investigating how other variables are related may offer further insights into what makes a wine good.

### Did you create any new variables from existing variables in the dataset?
Not yet, but I will group the quality ratings into three categories: bad, average and good.

### Of the features you investigated, were there any unusual distributions? Did you perform any operations on the data to tidy, adjust, or change the form of the data? If so, why did you do this?

The loaded dataset was already tidy.  

* There were 132 wines with 0 citric acid content which struck me as unusual. This is likely to be an error. 

* Quality, alcohol, sulphates, volatile acidity, citric acid (when the anomalous zero readings are discounted), density, pH and fixed acidity are approximately normally distributed. 
 
* The x-axes of the variables chlorides, residual sugar, total sulfur dioxide and free sulfur dioxide were log transformed. 


# Bivariate Plots Section

I will begin this section with a visualization of the correlation values between each feature in the data set, that can be used to guide and focus the following analysis. 


```{r echo=FALSE, Bivariate_Plots_1, fig.width = 12}
ggcorr(wine, label = TRUE, hjust = 0.9)
```

This shows the variables that correlate most highly with quality. These are:

* alcohol (positive correlation)

* volatile.acidity (negative correlation)

* sulphates (positive correlation)

* citric.acid (positive correlation)

* total.sulfur.dioxide (negative correlation)

* density (negative correlation)

It also reveals other correlations that might be worth investigating, such as :

* alcohol vs density is correlated as expected. It might be interesting to see how residual sugar affects that.
* fixed acidity and density are highly correlated. 
* chlorides and sulphates
* citric and fixed acid show a strong correlation with pH. This comes as no surprise. 
* There are strong correlations between volatile acidity and citric acid & fixed acidity and citric acid. This might be interesting to compare and contrast between wines of different qualities. 


Funnily, the quality is correlated, albeit only slightly with the number assigned to each wine. Perhaps the wine tasters got drunk during the tasting?

Let's now plot scatterplots of quality vs a number of chemical properties. To make the plots more informative, boxplots and trendlines are also included. Below each plot, the correlation coefficient is included. 

```{r echo=FALSE, Bivariate_Plots_2}

wine$quality.factor <- factor(wine$quality)

scatter_plot <- function(variable) {
  ggplot(aes_string(x = "quality.factor", y = variable), data = wine) +
    geom_point(alpha = 0.3, position = "jitter") +
    geom_boxplot(alpha = 0.5) + 
    stat_summary(fun.y = "mean",
               geom = "point",
               color = "red",
               shape =  8,
               size = 4) 
}

scatter_plot("alcohol") + 
  geom_smooth(aes(quality-2, alcohol), 
              data = wine, 
              method = "lm", 
              se = FALSE, 
              size = 1)
cor(wine$quality, wine$alcohol)

scatter_plot("volatile.acidity") + 
  geom_smooth(aes(quality-2, volatile.acidity), 
              data = wine, 
              method = "lm", 
              se = FALSE, 
              size = 1)
cor(wine$quality, wine$volatile.acidity)

scatter_plot("sulphates") + 
  geom_smooth(aes(quality-2, sulphates), 
              data = wine, 
              method = "lm", 
              se = FALSE, 
              size = 1)
cor(wine$quality, wine$sulphates)

scatter_plot("citric.acid") + 
  geom_smooth(aes(quality-2, citric.acid), 
              data = wine, 
              method = "lm", 
              se = FALSE, 
              size = 1)
cor(wine$quality, wine$citric.acid)

scatter_plot("total.sulfur.dioxide") + 
  geom_smooth(aes(quality-2, total.sulfur.dioxide), 
              data = wine, 
              method = "lm", 
              se = FALSE, 
              size = 1)
cor(wine$quality, wine$total.sulfur.dioxide)

# Correlation when we only look at wines with quality 5 or above.
# with(subset(wine, wine$quality >= 5), cor.test(quality, total.sulfur.dioxide))
```

```{r echo=FALSE, Bivariate_Plots_2b}
scatter_plot("density") + 
  geom_smooth(aes(quality-2, density), 
              data = wine, 
              method = "lm", 
              se = FALSE, 
              size = 1)
cor(wine$quality, wine$density)

scatter_plot("chlorides") +
  geom_smooth(aes(quality-2, chlorides), 
              data = wine, 
              method = "lm", 
              se = FALSE, 
              size = 1) + 
  scale_y_log10()
cor(wine$quality, wine$chlorides)

scatter_plot("pH") + 
  geom_smooth(aes(quality-2, pH), 
              data = wine, 
              method = "lm", 
              se = FALSE, 
              size = 1)
cor(wine$quality, wine$pH)

scatter_plot("free.sulfur.dioxide") + 
  geom_smooth(aes(quality-2, free.sulfur.dioxide), 
              data = wine, 
              method = "lm", 
              se = FALSE, 
              size = 1)
cor(wine$quality, wine$free.sulfur.dioxide)

scatter_plot("residual.sugar") + 
  geom_smooth(aes(quality-2, residual.sugar), 
              data = wine, 
              method = "lm", 
              se = FALSE, 
              size = 1)
cor(wine$quality, wine$residual.sugar)

scatter_plot("fixed.acidity") + 
  geom_smooth(aes(quality-2, fixed.acidity), 
              data = wine, 
              method = "lm", 
              se = FALSE, 
              size = 1)
cor(wine$quality, wine$fixed.acidity)

```

## Some Observations

* A trend of increasing alcohol content in higher quality wines is observed. Alcohol is the chemical property that correlates most with quality, with r = 0.47.

* Volatile acidity and quality are negatively correlated. We also see the IQR for volatile acidity decrease as the wine quality increases. 

* Sulphates and quality are correlated. There are many outliers for this distribution.

* Citric acid and quality are positively correlated. 

* The scatterplot for total sulfur dioxide is interesting because the boxplots for the bad wines (quality values of 3 and 4) are similar to those for the good wines (7 and 8). When we look at wines with quality of 5 or above, the correlation between total sulfur dioxide and quality has r = -0.25. This might be interesting to follow up on later in the investigation.

* The density varies very little from wine to wine. It is approximately equal to one, the density of water. There is a moderate negative correlation with good wines being, in general, "lighter".

* The scatterplot for chlorides supports what was observed on the histogram. The vast majority of values reside close to the median, and then the distribution has long tails. This is particularly true for wines with quality 5,6 and 7. Interestingly, the very good wines (rated 8) have much closer chloride values.

* The pH scatterplot show that good wines are slightly more acidic, but this relationship is weak with r = 0.06.

* The scatterplot for free sulfur dioxide is similar to that for total sulfur dioxide. We see a striking similarity between the free sulfur dioxide contents of wines rated 3 and 8.

* Residual sugar does not correlate with the quality of a wine. 

* There is a weak correlation between fixed acidity and quality. I'm surprised to see the similarity in the boxplots for wines rated 3 and wines rated 8. 

## Other Relationships

Let's begin by looking at density. The correlation coefficient in included beneath each figure. 

```{r echo=FALSE, Bivariate_Plots_3}
bivar_plot <- function (var1, var2) {
  ggplot(aes_string(var1, var2),
       data = wine) + 
  geom_point(alpha = .5) + 
  geom_smooth(method = "lm",
              se = FALSE, 
              size = 1)
}

bivar_plot("density", "fixed.acidity")
cor(wine$density, wine$fixed.acidity)

bivar_plot("density", "alcohol")
cor(wine$density, wine$alcohol)

bivar_plot("density", "residual.sugar")
cor(wine$density, wine$residual.sugar)

```

From the plots above we observe:

* fixed acidity and density are strongly and positively correlated with r = 0.67. 
* alcohol and density are negatively correlated with r = -0.50.

* residual sugar and density are moderately and positively correlated with r = 0.36.

* In the multivariate analysis, it will be interesting to add the quality rating as colour to these plots. 

Let's now look at a few other relationships.

```{r echo=FALSE, Bivariate_Plots_4}

bivar_plot("chlorides", "sulphates") 
cor(wine$chlorides, wine$sulphates)

bivar_plot("pH", "fixed.acidity")
cor(wine$pH, wine$fixed.acidity)

bivar_plot("pH", "volatile.acidity")
cor(wine$pH, wine$fixed.acidity)

```

Observations:

* The correlation matrix makes us think that chlorides and sulphates are strongly correlated. In fact, the scatterplot suggests that most wines have low sulphates and chlorides content. It is possible that a few outliers are creating a false correlation. 

* Fixed acidity and volatile acidity are correlated with pH.

# Bivariate Analysis

### Talk about some of the relationships you observed in this part of the investigation. How did the feature(s) of interest vary with other features in the dataset?

Many of the scatterplots above confirmed what was suspected from the correlation visualisation at the start of the section. We saw for example that alcohol and quality correlate strongly together and that there is a negative correlation between volatile acidity and quality. We saw some relationships that weren't expected. For example, the distribution for free and total sulfur dioxide were similar for bad wines (rated 3) and good wines (rated 8). 

### Did you observe any interesting relationships between the other features (not the main feature(s) of interest)?

We saw interesting relationships between density, fixed acidity, alcohol and residual sugar. We also saw that the correlation between chlorides and sulphates might be misleading and that pH and acidity were correlated, as expected.

### What was the strongest relationship you found?
The strongest relationship I found was between density and fixed acidity with r = 0.67. For the wine quality, alcohol level was the chemical property with the highest correlation with r = 0.47.


# Multivariate Plots Section

Begin by grouping all wines into 3 categories: 

* bad (rated 3 - 4)
* average (rated 5 - 6) and 
* good (rated 7 - 8)

```{r echo=FALSE, Multivariate_Plots_1}

wine$quality.rating <- ifelse(wine$quality <= 4, "bad", 
                              (ifelse(wine$quality <= 6, "average", "good"))
                              )

wine$quality.rating <- factor(wine$quality.rating, levels = c("bad", "average", "good"))

ggplot(wine, aes(y = volatile.acidity , x = alcohol, color = quality.rating)) +
  geom_point() + 
  scale_colour_brewer(palette="Reds") +
  theme_dark() +
  geom_smooth(method = "lm", 
              se = FALSE, 
              size = 1)

```

In the plot above, the regression lines nicely show the difference between the different quality wines. The good wines tend to have lower volatile acidity at all alcohol levels.

Let's do the same for sulphates.

```{r echo=FALSE, Multivariate_Plots_2}

ggplot(wine, aes(y = sulphates , x = alcohol, color = quality.rating)) +
  geom_point() + 
  scale_colour_brewer(palette="Reds") +
  theme_dark() +
  geom_smooth(method = "lm", 
              se = FALSE, 
              size = 1)

```

We see that for a given alcohol percentage, the amount of sulphates is higher. 

Let's plot fixed acidity and alcohol vs density and colour the scatter points by the quality rating. 

```{r echo=FALSE, Multivariate_Plots_3}
grid.arrange(
  ggplot(wine, aes(density, fixed.acidity, color = quality.rating)) + 
  geom_point(alpha = .5) + 
   scale_colour_brewer(palette = "Reds") + 
   theme_dark() +
  geom_smooth(method = "lm", 
              se = FALSE, 
              size = 1)
  ,
ggplot(wine, aes(density, alcohol, color = quality.rating)) + 
  geom_point(alpha = .5) + 
   scale_colour_brewer(palette = "Reds") + 
   theme_dark() +
  geom_smooth(method = "lm",
              se = FALSE, 
              size = 1)
)
```

Here we can see a clear distinction between good, average and bad wines. For a given density, good wines tend to have a higher fixed acidity content and higher alcohol. 

Let's see what the chlorides vs sulphates scatterplot looks like when we consider the quality of the wine. 

```{r echo=FALSE, Multivariate_Plots_4}
ggplot(wine, aes(chlorides, sulphates, color = quality.factor)) + 
  geom_point(alpha = .5) + 
   scale_colour_brewer(palette = "Reds") + 
   theme_dark()
```

Let's facet by quality rating and zoom in on the bulk of the data.

```{r echo=FALSE, Multivariate_Plots_5}
ggplot(wine, aes(chlorides, sulphates, color = quality.factor)) + 
  geom_point(alpha = .5) + 
   scale_colour_brewer(palette = "Reds") + 
   theme_dark() +
  xlim(0, quantile(wine$chlorides, 0.95)) + 
  ylim(0.25, quantile(wine$sulphates, 0.95)) + 
  facet_wrap(~ quality.rating)
```

Wines of all quality tend to have chloride values between 0.05 and 0.125 (although wines rated 8 tend to be between 0.05 and 0.0875). We see that the sulphates values shift up the better the wine. 


# Multivariate Analysis

### Talk about some of the relationships you observed in this part of the investigation. Were there features that strengthened each other in terms of looking at your feature(s) of interest?

The plots of volatile acidity vs alcohol and sulphates vs alcohol, both coloured by quality rating, didn't reveal much more than what the bivariate analysis showed. It only reinforced that good wines tend to have higher alcohol content, lower volatile acidity and higher sulphates. It was interesting plotting fixed acidity vs density and and alcohol vs density and seeing the datapoints group into bad, average and good wines. Interestingly, the trendlines for bad and average wines were almost coincident, indicating that other factors were responsible for their differentiation. 


------

# Final Plots and Summary

### Plot One
```{r echo=FALSE, Plot_One}
ggplot(aes(x = quality.factor, y = alcohol), data = wine) + 
       ggtitle("Alcohol Levels vs Wine Quality") +
  xlab("Quality Rating") +
  ylab("Alcohol (% by content)") + 
    geom_point(alpha = 0.3, 
               position = "jitter") +
    geom_boxplot(alpha = 0.5) + 
    stat_summary(fun.y = "mean",
               geom = "point",
               color = "red",
               shape =  8,
               size = 4) + 
  geom_smooth(aes(quality-2, alcohol), 
              data = wine, 
              method = "lm", 
              se = FALSE, 
              size = 1)

cor(wine$quality, wine$alcohol)
```

### Description One
My objective was to understand which chemical properties influence the quality of a given red wine. It is for this reason that I chose to display a scatterplot of alcohol content vs wine quality, as the alcohol level is the property that correlates highest with quality, with r = 0.47. We can see that very bad wines (rated 3) have a mean alcohol content of approximately 10% whereas very good wines (rated 8) have an alcohol content of approximately 12%.

### Plot Two
```{r echo=FALSE, Plot_Two}
ggplot(aes(x = quality.factor, y = volatile.acidity), data = wine) + 
       ggtitle("Volatile Acidity Levels vs Wine Quality") +
  xlab("Quality Rating") +
  ylab("Acetic Acid (g / dm^3)") + 
    geom_point(alpha = 0.3, 
               position = "jitter") +
    geom_boxplot(alpha = 0.5) + 
    stat_summary(fun.y = "mean",
               geom = "point",
               color = "red",
               shape =  8,
               size = 4) + 
  geom_smooth(aes(quality-2, volatile.acidity), 
              data = wine, 
              method = "lm", 
              se = FALSE, 
              size = 1)

cor(wine$quality, wine$volatile.acidity)
```

### Description Two
I have also included a similar plot for volatile acidity vs wine quality. I did this because I believe that this plot, along with plot 1, offers insight into what makes a good wine. It shows that the better the wine the lower its volatile acidity levels. It's a moderately strong correlation with r - -0.39.



### Plot Three
```{r echo=FALSE, Plot_Three}
ggplot(wine, aes(density, alcohol, color = quality.rating)) + 
  ggtitle("Alcohol Levels vs Density") +
  ylab("Alcohol (% by content)") +
  xlab("Density (g / cm^3)") +
  geom_point(alpha = .5) + 
   scale_colour_brewer(palette = "Reds") + 
   theme_dark() +
  geom_smooth(method = "lm", 
              se = FALSE, 
              size = 1)
```

### Description Three
My final plot illustrates an interesting relationship between other features that are not the main feature of interest of my investigation. It shows the relationship between density and alcohol content. For a given density, a good wine has higher alcohol content than a bad or average wine. 

------

# Reflection
The red wine data set contains information on 1599 Portugese Red Wines. To understand what chemical properites affect their quality, I started by understanding the individual variables in the data set, and seeing which ones correlated most highly with the quality rating. Based on these findings, I explored the data further, concentrating on the effect of alcohol, volatile acidity, sulphates content and citric acid.

My findings showed that most good red wines have a high alcohol, sulphate and citric acid levels and low volatile acidity. I consider this to be a success. I struggled to get a killer plot that would really help me understand what chemical properties made a good wine. Perhaps that is the learning - that there are many ways to skin a cat and there's no one way to make a good wine. 

Further work could be carried out to build a linear model to predict the quality of red wine. It would be interesting to incorporate variables that correlated less strongly with quality to refine the model. 

It would also be interesting to see the extent to which the conclusions I drew above apply to wines from other areas of the world. This analysis was performed on Portuguese "Vinho Verde" Red Wine. Would it also apply to a wine from California or Peru?